---
title: "Modeling Violent Crime in Philadelphia"
author: "Lallen Johnson"
date: "`r Sys.Date()`"
output: html_document
---

My objective is to model the relationship between indicators of poverty and violent crime across Philadelphia census tracts. To do this, I'm using the publicly available ProgressPHL [dataset](https://controller.phila.gov/philadelphia-audits/progressphl/#/). The data detail a collection of social, economic, and environmental indicators that fall within three thematic dimensions: Basic Human Needs, Foundations of Wellbeing, and Opportunity. 

```{r Importing Data}
setwd("C:/Users/johnsonl/OneDrive - american.edu/R/Philadelphia_Violent_Crime")
library(haven)
Philly <-read.csv("progressphl_data_download.csv", header = TRUE)
head(Philly)
```

```{r Cleaning Data}
library(reshape) #Using reshape package to change dataframe from long to wide format
phillyWide <- cast(Philly, geoid + neighborhood_name + region_name + tract_name ~ variable, value = "value")

library(janitor) #Using janitor package to replace spaces w/in variable names w/ underscores
phillyWide <- clean_names(phillyWide)
ls(phillyWide)
```

```{r Outcome Distribution}
library(summarytools)
descr(phillyWide$violent_crime_rate)
library(tidyverse)
histViol <- ggplot(data=phillyWide, aes(x=violent_crime_rate)) + geom_histogram(color="red", fill="white") + labs(x="Violent Crime Rate per 1,000", y="Count of Tracts")
histViol
```

```{r Variable Transformation}
phillyWide$ln_viol_cr <- log(phillyWide$violent_crime_rate)
histln_viol_cr <- ggplot(data=phillyWide, aes(x=ln_viol_cr)) + geom_histogram(color="black", fill="white") + labs(x="Logged Violent Crime Rate per 1,000", y="Count of Tracts")
histln_viol_cr
```

```{r Food Stamp Distribution}
descr(phillyWide$food_stamp_usage)
histFoodStamps <- ggplot(data=phillyWide, aes(x=food_stamp_usage)) + geom_histogram(color="blue", fill="white") + labs(x="% Households Reporting Food Stamp Usage", y="Count of Tracts")
histFoodStamps
```

```{r Residential Isolation Distribution}
descr(phillyWide$residential_isolation)
histResIsol <- ggplot(data=phillyWide, aes(x=residential_isolation)) + geom_histogram(color="black", fill="white") + labs(x="Residential Isolation Index")
histResIsol
```

```{r Bivariate Relationships}
scatterFSV <- ggplot(data = phillyWide, aes(x = food_stamp_usage, y = ln_viol_cr)) + geom_point() + geom_smooth(method = lm) + labs(x = "% Households Reporting Food Stamp Usage", y = "Logged Violent Crime Rate per 1,000")
scatterFSV

scatterRIV <- ggplot(data = phillyWide, aes(x=residential_isolation, y=ln_viol_cr)) + geom_point() + geom_smooth(method=lm) + labs(x="Residential Isolation Index", y="Logged Violent Crime Rate per 1,000")
scatterRIV
```

```{r Linear Regression 1}
summary(m1 <-lm(ln_viol_cr ~ food_stamp_usage, data=phillyWide))
#Inf values on logged violent crime rate prevent linear model from running
phillyWide[sapply(phillyWide, is.infinite)]<-NA
summary(m1 <-lm(ln_viol_cr ~ food_stamp_usage, data=phillyWide))
```

```{r Margins 1}
foodstamp <- data.frame(food_stamp_usage = c(20, 40, 80))
predict(m1, foodstamp)
```

```{r Linear Regression 2}
summary(m2 <-lm(ln_viol_cr ~ food_stamp_usage + residential_isolation, data=phillyWide))
```

```{r Linear Regression 3}
summary(m3 <-lm(ln_viol_cr ~ food_stamp_usage*residential_isolation, data=phillyWide))
```

```{r Margins 2}
library(sjPlot)
plot_model(m3, type="int", mdrt.values = "meansd", axis.title=c("% Households Reporting Food Stamp Usage","Logged Violent Crime Rate per 1,000"), title="Predictive Margins of Violent Crime in Philadelphia Tracts", legend.title="Residential Isolation")
```

#Still working on this